---
title: "Getting Started with rpolecat"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with rpolecat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The **rpolecat** package provides tools for downloading, managing, and analyzing
the POLECAT political event data from Harvard Dataverse. This vignette walks
through a complete workflow, from installation through analysis.

## Installation

Install rpolecat from GitHub using the **remotes** package:

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("andybega/rpolecat")
```

```{r load-library}
library(rpolecat)
```

## Setup

### Dataverse API token

POLECAT data is hosted on Harvard Dataverse. To download it, you need a
Dataverse API token. You can obtain one by creating a free account at
<https://dataverse.harvard.edu> and generating a token from your account
settings.

Set the token as an environment variable in your `.Renviron` file:

```{r set-token, eval = FALSE}
# Open .Renviron for editing
usethis::edit_r_environ()

# Add this line (replace with your actual token):
# DATAVERSE_KEY=your-api-token-here
```

You can verify that your token is set correctly:

```{r check-token, eval = FALSE}
dataverse_api_token()
```

### Configure the package

Use `setup_polecat()` to tell rpolecat where to store data files locally. This
avoids having to pass the directory path to every function call.

```{r setup-polecat, eval = FALSE}
# Set the local data directory for this session
setup_polecat("~/data/polecat")

# Persist across sessions by writing to .Rprofile
setup_polecat("~/data/polecat", r_profile = TRUE)
```

You can check or change your settings at any time:

```{r get-opts, eval = FALSE}
get_polecat_opts()
```

## Downloading data

Use `download_polecat()` to download POLECAT data from Harvard Dataverse to
your local machine:

```{r download-all, eval = FALSE}
# Download all available data
download_polecat("~/data/polecat")
```

You can also download data for specific years, or skip files you already have:

```{r download-selective, eval = FALSE}
# Download only 2024 data
download_polecat("~/data/polecat", years = 2024)

# Skip files already on disk (useful for updates)
download_polecat("~/data/polecat", skip_existing = TRUE)

# Preview what would be downloaded without actually downloading
download_polecat("~/data/polecat", dryrun = TRUE)
```

## Reading data

Once downloaded, read the data into R with `read_polecat()`:

```{r read-data, eval = FALSE}
# Read all local data
events <- read_polecat("~/data/polecat")

# Read only specific years
events_2024 <- read_polecat("~/data/polecat", years = 2024)
```

## Exploring included data

The package ships with several built-in datasets that are useful for exploring
and understanding POLECAT without needing to download the full data.

### Sample data

`polecat_sample` contains a small subset of POLECAT events for demonstration
purposes:

```{r polecat-sample}
data(polecat_sample)
str(polecat_sample)
```

### Reference tables

The package includes several reference tables describing the POLECAT coding
scheme:

```{r types}
# Event types (CAMEO-based)
data(types)
head(types)
```

```{r modes}
# Event modes
data(modes)
head(modes)
```

```{r contexts}
# Event contexts
data(contexts)
head(contexts)
```

```{r quad-categories}
# Quad category definitions
data(quad_categories)
quad_categories
```

## Data cleaning

### Normalize

`normalize_polecat()` standardizes a raw POLECAT data frame by parsing date
columns, adding a year variable, and ensuring consistent column types:

```{r normalize, eval = FALSE}
events <- read_polecat("~/data/polecat")
events <- normalize_polecat(events)
```

### Deduplicate

The POLECAT data may contain duplicate events. Use `deduplicate_polecat()` to
remove them:

```{r deduplicate}
nrow(polecat_sample)
deduped <- deduplicate_polecat(polecat_sample)
nrow(deduped)
```

## Filtering

rpolecat provides several functions for filtering events.

### General filtering

`filter_polecat()` is a flexible function for subsetting events by country,
date, event type, and more:

```{r filter-general}
# Filter to events involving the United States
us_events <- filter_polecat(polecat_sample, countries = "USA")
nrow(us_events)

# Filter by source and target countries
us_to_chn <- filter_polecat(polecat_sample,
                             src_countries = "USA",
                             tgt_countries = "CHN")
nrow(us_to_chn)
```

### Dyadic, intrastate, and interstate filters

These convenience functions filter events based on the relationship between
source and target actors:

```{r filter-dyadic}
# Keep only events with both a source and target country
dyadic <- filter_dyadic(polecat_sample)
nrow(dyadic)
```

```{r filter-intrastate}
# Keep only events where source and target are in the same country
intra <- filter_intrastate(polecat_sample)
nrow(intra)
```

```{r filter-interstate}
# Keep only events where source and target are in different countries
inter <- filter_interstate(polecat_sample)
nrow(inter)
```

## Aggregation

`aggregate_polecat()` rolls up event-level data to common units of analysis
used in political science research.

### Units of analysis

The `unit` argument controls the level of aggregation:

```{r agg-country-month}
# Country-month event counts (the default)
cm <- aggregate_polecat(polecat_sample, unit = "country-month")
head(cm)
```

```{r agg-country-year}
# Country-year
cy <- aggregate_polecat(polecat_sample, unit = "country-year")
head(cy)
```

```{r agg-dyad-month}
# Dyad-month (source-target country pairs)
dm <- aggregate_polecat(polecat_sample, unit = "dyad-month")
head(dm)
```

### Measures

The `measure` argument controls what summary statistic is computed:

```{r agg-goldstein}
# Mean Goldstein score per country-month
gs <- aggregate_polecat(polecat_sample,
                        unit = "country-month",
                        measure = "goldstein")
head(gs)
```

### Quad category breakdowns

Set `by_quad = TRUE` to get separate counts for each quad category (Verbal
Cooperation, Material Cooperation, Verbal Conflict, Material Conflict):

```{r agg-quad}
quad_counts <- aggregate_polecat(polecat_sample,
                                 unit = "country-month",
                                 by_quad = TRUE)
head(quad_counts)
```

## Quad classification

POLECAT events can be mapped to the four standard quad categories based on
their event type codes.

### Classify events

`classify_quad()` adds a quad category column to a POLECAT data frame:

```{r classify-quad}
classified <- classify_quad(polecat_sample)
table(classified$quad_category)
```

### Look up quad categories

`get_quad()` returns the quad category for one or more event type codes:

```{r get-quad}
get_quad("01")
get_quad(c("01", "10", "14", "19"))
```

## Country codes

POLECAT uses ISO 3166-1 alpha-3 country codes. The package provides functions
to convert these to the Gleditsch-Ward (GW) and Correlates of War (COW) numeric
codes commonly used in political science datasets.

These functions require the **countrycode** package.

```{r gwcode, eval = FALSE}
# Convert to Gleditsch-Ward codes
events <- polecat_to_gwcode(polecat_sample)
head(events[, c("src_actor", "src_gwcode")])
```

```{r cowcode, eval = FALSE}
# Convert to Correlates of War codes
events <- polecat_to_cowcode(polecat_sample)
head(events[, c("src_actor", "src_cowcode")])
```

## Database support

For working with the full POLECAT data (which can be very large), rpolecat
supports storing events in a DuckDB or SQLite database and querying them
directly.

### Create a database

```{r create-db, eval = FALSE}
# Create a DuckDB database from local data files
create_polecat_db("~/data/polecat/polecat.duckdb",
                  data_dir = "~/data/polecat")
```

### Query the database

```{r query-db, eval = FALSE}
# Run SQL queries against the database
result <- query_polecat(
  "SELECT src_actor, COUNT(*) AS n FROM polecat WHERE year = 2024 GROUP BY src_actor ORDER BY n DESC LIMIT 10",
  db_path = "~/data/polecat/polecat.duckdb"
)
head(result)
```

## Managing data

rpolecat includes utilities for keeping your local data up to date and managing
disk space.

### Update data

`update_polecat()` checks for new or updated files on Dataverse and downloads
them:
```{r update, eval = FALSE}
update_polecat()
```

### Compress data

`compress_polecat()` compresses (or decompresses) local data files to save disk
space:

```{r compress, eval = FALSE}
# Compress all local files
compress_polecat()

# Decompress
compress_polecat(action = "decompress")
```

### List files

`list_polecat_files()` shows the local data files along with their sizes:

```{r list-files, eval = FALSE}
list_polecat_files()
```

You can also list files available on Dataverse:

```{r list-dv-files, eval = FALSE}
list_dataverse_files()
```
